\documentclass[11pt]{article}

% basic packages
\usepackage[margin=1in]{geometry}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{william}
\newcommand{\fd}[2]{\frac{\delta #1}{\delta #2}} % functional derivative
\usepackage{tikz-cd}

% page formatting
\usepackage{fancyhdr}
\pagestyle{fancy}

\renewcommand{\sectionmark}[1]{\markright{\textsf{\arabic{section}. #1}}}
\renewcommand{\subsectionmark}[1]{}
\lhead{\textbf{\thepage} \ \ \nouppercase{\rightmark}}
\chead{}
\rhead{}
\lfoot{}
\cfoot{}
\rfoot{}
\setlength{\headheight}{14pt}

\linespread{1.03} % give a little extra room
\setlength{\parindent}{0.2in} % reduce paragraph indent a bit
\setcounter{secnumdepth}{2} % no numbered subsubsections
\setcounter{tocdepth}{2} % no subsubsections in ToC

\begin{document}

% make title page
\thispagestyle{empty}
\bigskip \
\vspace{0.1cm}

\begin{center}
{\fontsize{22}{22} \selectfont Lecture Notes on}
\vskip 16pt
{\fontsize{36}{36} \selectfont \bf \sffamily QFT}
\vskip 24pt
{\fontsize{18}{18} \selectfont \rmfamily Will Lancer} 
\vskip 6pt
{\fontsize{14}{14} \selectfont \ttfamily will.m.lancer@gmail.com} 
\vskip 24pt
\end{center}

{\parindent0pt \baselineskip=15.5pt}
\noin
Resources used:
\begin{itemize}
    \item Xi Yin's 253ab courses.
    \item Steven Weinberg's two volumes.
    \item Daniel Harlow's notes.
    \item Sidney Coleman's QFT. 
    \item Peskin and Schroeder.
    \item Schwartz.
    \item Srednicki.
    \item \emph{FIELDS} by Warren Siegel.
\end{itemize}
We mostly follow Weinberg's notation, except where his notation sucks.

% make table of contents
\newpage
\microtoc
\newpage

% main content

\section{Overview}

\note{overview things here; talk about how single-particle rel
qm is inconsistent; talk about how particles are fundamental,
and not fields (not so sure about this anymore)}


Which is more fundamental, particles or fields? Most physicists
are likely to say fields---they are \emph{wrong}: particles are fundamental.
To see this, just ask an experimentalist what they measure. They measure
\emph{particles}! In the calorimeter at the LHC, \emph{particles} hit the
walls of the collider, blazing paths through the detector that we can reconstruct
as paths these particles take. The energy and momentum of these particles, combined
with the ingenuity of experimentalists, tells you exactly what kind of particle
this was. There is no mention of fields anywhere in this picture. \note{can
steelman this, but should also fight back}\\


\note{think about this more}\\
An ``axiom'' of QFT is that it has local, microcausal field operators. These
represent local ``measurement'' on our system. One can think of this as taking
inspiration from electromagnetism, since that's our only example of a truly
local observable (e.g. the electric field; you measure it at a \emph{point}.
This is obvious actually, as this is what \emph{must} occur mathematically,
i.e. $\phi \colon M \to \reals$, so there is no sense of asking its value at
anywhere but a point you ``can reach''). Since these are, by definition, observables,
they also must be Hermitian. Since things are local, they must obey Poincaré
symmetry. Enumerating these conditions for the scalar field gives
\begin{enumerate}
    \item \textbf{Poincaré symmetry}: under a Poincaré transformation $U(\Lambda, a)$,
    our fields must transform as
    \begin{align*}
        U(\Lambda, a) \phi(x) U(\Lambda, a)^{-1} = \phi(\Lambda x + a).
    \end{align*}
    \item \textbf{Microcausality}: our fields must be local and causal, e.g. spacelike
    separated measurements can't effect each other. This corresponds to
    \begin{align*}
        & [\phi(x), \phi(y)] = 0 
        & (x - y)^2 > 0.
    \end{align*}
    \item \textbf{Unitarity}: as stated, the fields must be Hermitian,
    $\phi = \phi^\dagger$.
\end{enumerate}


\section{Relativistic Quantum Mechanics}

We start by recalling the axioms of quantum mechanics\footnote{
    Of course, there are different axioms than this that give
    identical results. ``Axioms'' is really a misnomer; they're
    just assumptions that let us build the rest of the theory.
    There is nothing intrinsically fundamental about this set of
    axioms as opposed to other ones.
}:


\begin{iidea}
    [Axioms of quantum mechanics]
    The axioms of quantum mechanics are:
    \begin{itemize}
        \item Quantum mechanical \vocab{states} are represented by \vocab{rays}
        in a \vocab{Hilbert space}. We will denote Hilbert spaces by $\mathcal{H}$.
        \item \vocab{Observables} are represented by self-adjoint operators on
        our Hilbert space. Observed quantities are the eigenvalues of vectors in $\mathcal{H}$.
        \item Let $\mathcal{R}$ be a ray. If $\mathcal{R}_1, \ldots, \mathcal{R}_n$ are a
        set of mutually orthogonal rays, the probabilty of observing $\mathcal{R}$ in
        $\mathcal{R}_i$ is
        \begin{align*}
            \mathbb{P}(\mathcal{R} \to \mathcal{R}_i) = | \braket{\psi}{\phi_i} |^2,
        \end{align*}
        where $\psi \in \mathcal{R}$ and $\phi_i \in \mathcal{R}_i$.
    \end{itemize}
\end{iidea}

\noin
Remarks:

\begin{itemize}
    \item Recall that rays are equivalence classes of kets up to
    phase and normalization. Since this is an equivalence class,
    WLOG we can choose our representative to be normalized.
    \item Self-adjoint or Hermitian? \note{comment on this}
    \item In our third axiom, we make reference to experiment.
    How do we physically represent an experiment on our Hilbert space?
    \note{good question}
\end{itemize}

The most important kinds of operators are operators that realize
\emph{symmetry transformations} on our Hilbert space.

\note{finish}

\subsection{Symmetries}

\note{add the normal ones, and then ps
in the next section}

\note{for whatever reason}, this continuity may also be found in the
\emph{representation} of the symmetry.

\subsection{Poincaré symmetry and the little group}
Since we are doing relativistic quantum mechanics, let's
see how the relativistic symmetry group acts on our Hilbert space.
The Poincaré group in mostly-plus is given by
\begin{align*}
    \mathcal{P} = \reals^D \sdprod SO(D - 1, 1).
\end{align*}
Consider some Poincaré transformation $(\Lambda\indices{^\mu_\nu}, a^\mu)$.
By Wigner's theorem, this is represented as a linear unitary operator on
our Hilbert space; the group property is
\begin{align*}
    U(\Lambda, a) U(\bar{\Lambda}, \bar{a}) = U(\Lambda \bar{\Lambda}, \Lambda \bar{a} + a).
\end{align*}
Notice that this form implies that the Poincaré group is
a \emph{semi-direct product}, not a direct product (as the
translation entry is ``twisted'' by $\Lambda$). Since the Poincaré 
group is a Lie group, things are continuous.
Taylor-expanding $U(\Lambda, a)$ gives
\begin{align*}
    U(\Lambda, a) \simeq \delta^{\mu}_{\nu} - i a_\mu \mathcal{P}^\mu + \frac{i}{2} \omega_{\mu \nu} J^{\mu \nu}.
\end{align*}
The generators of $U(\Lambda, a)$ are very special: they are the
four-momentum operator and the angular momentum/boost operator.
Considering their components gives
\begin{align*}
    & P^\mu = (H, \mathbf{P}),
    & J^{\mu \nu} = \begin{cases}
        J^{0i} = K^i\\
        J^{ij} = \epsilon^{ijk}J_k,
    \end{cases}
\end{align*}
where $K^i$ is the generator of boosts.

We would now like to see how $U(\Lambda, a)$ acts on our states.
Choose our states to be labeled as $\ket{\bfk, \sigma}$, where
$\bfk$ is our spatial momentum and $\sigma$ is some set of internal
degrees of freedom (which have not yet been specified). Note that
we may well use $\ket{k, \sigma}$ instead of $\ket{\bfk, \sigma}$,
but we choose $\ket{\bfk, \sigma}$ here because it emphasizes that these
particles are all on-shell (i.e., that $k^0 = \sqrt{|\bfk|^2 + m^2}$).
First, note that this basis diagonalizes translations,
\begin{align*}
    U(0, a) \ket{\bfk, \sigma} = e^{- i \mathcal{P} a} \ket{\bfk, \sigma} = e^{-i k a} \ket{\bfk, \sigma}.
\end{align*}
Indeed, this is one of the reasons we chose our states to be labeled
by momentum eigenvalues; momentum is symmetric under translations. Let us
now study how Lorentz transformations act on these states. \emph{Define} an 
arbitrary state as
\begin{align*}
    \ket{\bfk, \sigma} \coloneqq \mathcal{N}(\bfk) U(L(\bfk)) \ket{\bfk_R, \sigma},
\end{align*}
where $\bfk_R$ is a given \vocab{reference momentum}\footnote{This will
be specific to the kinds of particles you are talking about, e.g.
$k_R = (E, 0, 0, E)$ is a reference momentum for the massless little
group in $D = 4$.}, $L(\bfk)$ is some Lorentz transformation 
that takes $\bfk_R \xrightarrow{L(\bfk)} \bfk$, and $\mathcal{N}(\bfk)$
a normalization factor to be determined. Consider $U(\Lambda) \ket{\bfk, \sigma}$
for arbitrary $\Lambda$. We have
\begin{align*}
    U(\Lambda) \ket{\bfk, \sigma} & = \mathcal{N}(\bfk) U(\Lambda) U(L(\bfk)) \ket{\bfk_R, \sigma}\\
    & = \mathcal{N}(\bfk) \mathcal{N}(\boldsymbol{\Lambda k})^{-1} U(L(\boldsymbol{\Lambda k})) U(\underbrace{L(\boldsymbol{\Lambda k})^{-1} \Lambda L(\bfk))}_{W}) \ket{\bfk_R, \sigma}.
\end{align*}
Let's study $W$. The following diagram commutes:
\[
    \begin{tikzcd}[column sep=large, row sep=large]
    \bfk_R \arrow[r, "\,L(\bfk)\,"] \arrow[rrd, "\,W\,"'] & \bfk \arrow[r, "\,\Lambda\,"] & \boldsymbol{\Lambda \bfk} \arrow[d, "\,L(\boldsymbol{\Lambda \bfk})^{-1}\,"] \\
    & & \bfk_R
    \end{tikzcd}
\]
Thus, $W$ fixes $\bfk_R$. We call $W$ a \vocab{Wigner rotation},
and we call the set of such $W$'s the \vocab{little group} for
our given $\bfk_R$. Note that the little group is specific
to the kind of reference momentum $\bfk_R$ is, e.g. massless
particles in $D = 4$ form one little group. 

The key observation to make is that while $U(W)$
can't change $\bfk_R$, it can change our $\sigma$
labels; thus, the $\sigma$ labels tell you
how the particle transforms under the \emph{little group}. 
We may generically write such a representation of the little group
as
\begin{align*}
    U(W) \ket{\bfk_R, \sigma} = \sum_{\sigma'} D_{\sigma \sigma'}(W) \ket{\bfk_R, \sigma'}.
\end{align*}
Notice that $D(W_1 W_2) = D(W_2) D(W_1)$, as we can see from the definition.
We now determine the normalization from orthogonality. Consider \note{finish}
Putting this all together, we have that
\begin{align*}
    \boxed{U(\Lambda)\ket{\bfk, \sigma} = \sqrt{\frac{\omega_{\boldsymbol{\Lambda \bfk}}}{\omega_{\bfk}}}\sum_{\sigma \sigma'} D_{\sigma \sigma'}(W(\bfk)) \ket{\boldsymbol{\Lambda \bfk}, \sigma'}.}
\end{align*}
\note{why does this hold for multiparticle states? I think because they
are diff vectors, so doesn't matter; can use raising and lowering ops i think}
\begin{align*}
    \boxed{U(\Lambda, a)\ket{\bfk_1, \sigma_1; \bfk_2, \sigma_2; \ldots ; \bfk_n, \sigma_n}
    = \prod_{i = 1}^n e^{-i k_i a}\sqrt{\frac{\omega_{\boldsymbol{\Lambda \bfk}_i}}{\omega_{\bfk_i}}} \sum_{\sigma_i \sigma_i'} D_{\sigma_i \sigma_i'}(W(\bfk_i)) \ket{\boldsymbol{\Lambda \bfk}_1, \sigma_1'; \boldsymbol{\Lambda \bfk}_2, \sigma_2'; \ldots, \boldsymbol{\Lambda \bfk}_n, \sigma_n'}}
\end{align*}
This will be useful when we talk about scattering later.

We now take a closer look at the $D$ matrices. \note{finish}

\subsection{Lagrangian QM}

Now we'll get into Lagrangian QM, which naturally introduces
the path integral and accompanying ideas (regularization, renormalization,
etc.).

\begin{iidea}
    [The path integral]
    Let $\bra{q_f}$ and $\ket{q_i}$ respectively be final
    and initial position eigenstates. Call the time-translation
    operator for a time $T$, $U(T)$. \note{finish}
\end{iidea}

\note{add lagrangian qm; reg, renorm, path integral derivation;
add exercises from yin and weinberg here too}

\section{Classical Field Theory}

We briefly review classical field theory here; for a more in-depth analysis,
see the \note{notes} on classical field theory.

As always, we start with an action:
\begin{align*}
    S[\phi] = \int d^D x \, \el = \int dt \int d^{D - 1}\mathbf{x} \, \el(\phi, \partial \phi, \ldots),
\end{align*}
where the bounds on this integral are generically over all of $\reals^D$,
but may be restricted to a subset of it (e.g. the upper half plane of $\reals^2$).
The equations of motion are derived from the action by demanding stationarity
\begin{align*}
    \fd{S}{\phi} = 0 \iff \boxed{\partial_\mu \left( \pd{\el}{(\partial_\mu \phi)} \right) - \pd{\el}{\phi} = 0.}
\end{align*}
If we interchange $\phi$ for a tensor field, e.g. $A_\mu$ or $h_{\mu \nu}$,
the ELE above become tensor equations: one for each component of the tensor
field. 

\note{talk about canonical coordinates here}


We now investigate symmetries of the action---we consider the infinitesimal
case and build from there. Let
\begin{align*}
    \phi \to \phi'(x) = \phi(x) + (\delta_\epsilon \phi)(x),
\end{align*}
where $(\delta_\epsilon \phi)(x) \coloneqq \epsilon \, \delta \phi(x)$,
with $\epsilon \ll 1$ and $\delta \phi(x)$ a field variation.
Notice only the field variable is primed. If we prime both
variables, we get an equality: $\phi'(x') = \phi(x)$. This
is fundamentally because the transformed field at the transformed point
should give the same measurement as the original field at the
original point. If the coordinate transformation above 
is a symmetry, that means that the action is unchanged
under such a transformation. Consider a non-symmetry
transformation:
\begin{align*}
    \phi \to \phi'(x) = \phi(x) + \tilde{\delta}_\epsilon \phi(x),
\end{align*}
where $(\tilde{\delta}_\epsilon \phi)(x) \coloneqq \epsilon(x) \delta \phi(x)$.
We call this process of making $\epsilon$ $x$-dependent ``gauging the symmetry''---we
also say this is turning a global symmetry into a local one. The action
should no longer be invariant under such a transformation, but should take the
form
\begin{align*}
    \tilde{\delta} S = \int d^D x \, \partial_\mu \epsilon(x) j^\mu(x) \;
    \xrightarrow{\epsilon(x) \, = \, \epsilon} \; \delta S = 0.
\end{align*}
Note that this implies that $\partial_\mu j^\mu = 0$ when $\epsilon$ is constant,
thus giving us a locally conserved current. Integrating $\tilde{\delta}S$ by parts 
gives an expression for $j^\mu(x)$,
\begin{align*}
    \boxed{j^\mu(x) = \pd{\el}{(\partial_\mu \phi)} \delta \phi - \mathcal{U},}
\end{align*}
where $\mathcal{U}$ is what's left over. Concretely,
\begin{align*}
    \tilde{\delta} S[\phi] 
    & = \int d^D x \, \pd{\el}{\phi} \, (\tilde{\delta}_\epsilon \phi)(x) + \pd{\el}{(\partial_\mu \phi)} \, \partial_\mu (\tilde{\delta}_\epsilon \phi)(x)\\
    & = \int d^D x \, \partial_\mu \epsilon(x) \left( \pd{\el}{(\partial_\mu \phi)} \delta \phi(x) \right) + \epsilon(x) \underbrace{\left(\pd{\el}{\phi} \delta \phi(x) + \pd{\el}{(\partial_\mu \phi)} \partial_\mu \delta \phi(x) \right)}_{\partial_\mu \, \mathcal{U}(x)}.
\end{align*}
We may also form a conserved charge from $j^\mu$ by integrating its
time component over some fixed time-slice. We have
\begin{align*}
    Q \coloneqq \int d^{D - 1} \bfx \, j^0(\bfx).
\end{align*}
One can see that $Q$ is conserved by taking its time derivative
and using the fact that $\partial_\mu j^\mu = 0$. One can also
check that $Q$ generates the symmetry $\delta \phi$ in the sense 
of Hamiltonian mechanics:
\begin{align*}
    \delta \phi(x) = \{ Q, \phi \}|_{\pi(\bfx) = \partial \el / \partial \dot{\phi}(\bfx)}
\end{align*}

The most important Noether current in a classical field theory is
the stress-energy-momentum tensor $j^\mu(x) = a^\nu T^\mu{}_\nu(x)$.
Consider its derivation: if we translate the fields by an infinitesimal
amount $a^\mu$, we have that
\begin{align*}
    \phi'(x') & = \phi(x)\\
    \phi'(x + a) & = \phi(x)\\
    \phi'(x) + a^\mu \partial_\mu \phi'(x) & = \phi(x)\\
    \implies \phi'(x) & = \phi(x) - a^\mu \partial_\mu \phi(x).
\end{align*}
So, $\epsilon \rightsquigarrow a^\mu$ and $\delta \phi \rightsquigarrow \partial_\mu \phi(x)$.
Note an important step: we replaced $\partial_\mu \phi'(x)$ with $\partial_\mu \phi(x)$.
We can do this because $\phi'(x) = \phi(x)$ to first-order: concretely,
\begin{align*}
    a^\mu \partial_\mu \phi'(x) - a^\mu \partial_\mu \phi(x)
    = a^\mu \partial_\mu (\phi'(x) - \phi(x))
    \simeq a^\mu \partial_\mu (\delta_a \phi(x)) \sim O(a^2),
\end{align*}
and thus we're safe to replace $\partial_\mu \phi'(x)$ with $\partial_\mu \phi(x)$.
Inputting these variations into the expression for $j^\mu(x)$ gives
\begin{align*}
    chung
\end{align*}

\begin{eexample}
    [Poincaré symmetry]
    \note{todo}
\end{eexample}






\section{Spin-0 QFTs}

\section{Scattering}

We now get into scattering theory. This is probably the most
important part of QFT, so we'll go relatively deep into it.

\subsection{The basics}

Composite particles decay, and fundamental particles \note{finish}

\subsection{Symmetries of the $S$-matrix}

\subsection{The LSZ reduction}

\begin{reemark}
    [On our derivation]
    For the LSZ reduction, we follow Haag and Ruelle by way of Yin.
    You may see his lecture notes on this from 253a linked on the
    first page of these notes.

    We will purposefully state many approximate results. One can get true equalities
    here, but the derivation is much more obtuse, so we will stick to our wave
    packets. Also, everything is basically the same anyways, so I don't care.
\end{reemark}

The goal of the LSZ reduction is to write our in and out
states in terms of our field operators, thus giving us a
connection between the $S$-matrix and perturbation theory.
A priori, there is no connection here to speak about---indeed, 
the LSZ reduction is \textbf{the most important
result in all of quantum field theory}.

We begin by defining a ``smeared field operator'' via
\begin{align*}
    \phi_f \coloneqq \int d^D x \, f(x) \phi(x).
\end{align*}
We assume our smearing function's Fourier transform is supported 
near the mass shell:
\begin{align*}
    \tilde{f}(k) \neq 0 \xleftrightarrow{\: \: \: \: \approx \: \: \: \:} k^2 + m^2 = 0.
\end{align*}
We must be able to talk about ``moving fields''---how else
are they supposed to travel out to infinity?
Consider the following transformation on $f$:
\begin{align*}
    f(k) \mapsto f^{(T)}(k) \text{ by } f^{(T)}(k) = e^{i(k^0 - \omega_\bfk)T} f(k).
\end{align*}
How does this transformation act on our field operators?
\note{add wave equation part here}
Furthermore, the support of $\tilde{f}$ kills the
multi-particle contribution of $\phi_{f} \ket{\Omega}$.
Recall
\begin{align*}
    \phi(x) \ket{\Omega} & = \int \dbar^{D - 1} \bfk \, e^{-ikx}\, \mathcal{Z}^{\rm eff}_{\bfk} \ket{\bfk}
    + \int_{\rm m.p.} d\alpha \, e^{-ip_\alpha x} \, \mathcal{Z}^{\rm eff}_{\alpha} \ket{\alpha}
\end{align*}
Applying the definition of $\phi_f$ and taking the inverse 
Fourier transform of $f$ gives
\begin{align*}
    \phi_f \ket{\Omega} & = \int d^D x \, \int \dbar^{D - 1} \bfk \, e^{-ikx}\, \mathcal{Z}^{\rm eff}_{\bfk} \ket{\bfk}
    + \int d^D x \int_{\rm m.p.} d\alpha \, e^{-ip_\alpha x} \, \mathcal{Z}^{\rm eff}_{\alpha} \ket{\alpha}\\
    & \approx \int \dbar^{D - 1} \bfk \, \tilde{f}(\bfk, \omega_\bfk)\, \mathcal{Z}^{\rm eff}_{\bfk} \ket{\bfk} + 0
\end{align*}
\note{explain why T transformation doesn't do anything}Thus,
\begin{align*}
    \boxed{\phi_f \ket{\Omega} \approx \phi_{f^{(T)}} \ket{\Omega}.}
\end{align*}
The intuition behind this is that ``there are a lot more operators than
states''. Thus we may define our states as
\begin{align*}
    \ket{\phi_{f_1}, \ldots, \phi_{f_n}} \coloneqq \phi_{f_n} \cdots \phi_{f_1} \ket{\Omega}.
\end{align*}
Thus, by taking inner products of our states, we may get the $S$-matrix
in terms of field operators acting on the vaccuum. We have \note{add $T$
variables on all of these guys below}

\subsection{Analyticity properties of the $S$-matrix}

\note{my goat xi yin}

\note{unitarity bounds too?}

\section{Spin-$1/2$ QFTs}



\section{Spin-$1$ QFTs}

\note{motivation}

\subsection{Gauge reundancy; $2 \neq 3$}

There are two cases for spin-$1$ particles: massless or massive.
They correspond to different representations of the little group,
respectively being $SO(D - 1)$ and $SO(D - 2)$. In $D = 4$, this means
that a massive spin-$1$ particle has \emph{three} spin DOF,
while a massless spin-$1$ particle has \emph{two} spin DOF.
This discontinuous jump in the number of DOF between massive and massless
is the root cause behind the difficulties of dealing with massless spin-$1$
particles.

Concretely, we describe the potential of (say) a photon using
a four-vector to manifest Poincaré symmetry, i.e.
\begin{align*}
    A_\mu(x) = \epsilon_\mu e^{ikx}.
\end{align*}
We can demand transversality, $k \cdot \epsilon = 0$, which takes us
from $4 \to 3$ DOF. If we have a massive particle, hooray! $3 = 3$.
But if we have a massless particle, we're dead in the water---there are no
other Lorentz invariant constraints that we can impose on $\epsilon$.
Thus, we have introduced \emph{gauge redundancy}: concretely, we
have defined the equivalence relation
\begin{align*}
    \epsilon_\mu \sim \epsilon_\mu + \alpha k_\mu.
\end{align*}
Thus, saying ``the'' polarization vector when referring to a photon
is a misnomer---you really mean ``a'' polarization vector. We can
see from this example that the biggest qualitative difference
between massless and massive spin-$1$ QFTs is the number of DOF
of their particles. This will cause lots of problems for us later,
so buckle in and let's go!

\note{general Lorentz transformation properties of spin-1 particles
go here;}



\note{\begin{itemize}
    \item particles; massive vs massless
    \item little group representations
    \item particles -> fields; massless vs massive
    \item Path integral construction; gauge fixing
\end{itemize}}


\section{Spin-$3/2$ QFTs}

\note{susy$\ldots$}

\section{Spin-$2$ QFTs}

\note{gravity!}

\note{maybe I should find a better organization scheme for this}

\section{Useful formulae}

\subsection{Integrals and special functions}

\note{schwinger parameters, feynman parameters, equivalence of the two}

\subsubsection{Schwinger Parameters}

We have that
\begin{align*}
    \int_0^\infty ds \, s^{- a - 1} e^{-bs - c/s} = 2 \left( \frac{b}{c} \right)^{1/2} K_a(2 \sqrt{bc}).
\end{align*}

\subsection{Dirac matrices}

We reproduce some of the results covered in the notes on
\note{group theory} here. We may form the Dirac matrices
in any dimension. The most useful is $D = 4$. A choice
of Dirac matrices in $D = 3 + 1$ is
\begin{align*}
    & \gamma^0 = - i \begin{pmatrix}
        0 & I_2\\
        - I_2 & 0
    \end{pmatrix}
    & \gamma^a = -i \begin{pmatrix}
        0 & \sigma^a\\
        - \sigma^a & 0
    \end{pmatrix}.
\end{align*}
You can construct the Dirac matrices in any dimension by
simply finding matrices that obey the Clifford algebra,
$\{ \gamma^\mu, \gamma^\nu \} = 2 \eta^{\mu \nu}$. Practically,
this is done by bootstrapping your way up from a given dimension
via the following iterative process:
\begin{itemize}
    \item \textbf{Even $\to$ odd}: to go from even to odd dimensions,
    add the chirality matrix to your basis, $\gamma_c = \alpha \gamma^1 \cdots \gamma^n$.
    Fix $\alpha$ by demanding $(\gamma_c)^2 = 1$. This gives two inequivalent
    irreps, $\{ \gamma^{(n)}, \gamma_c \}$ and $\{ \gamma^{(n)}, - \gamma_c \}$.
    \item \textbf{Odd $\to$ even}: to go from odd to even dimensions,
    tensor the existing representation by one Pauli matrix, and
    add the identity times another Pauli matrix to get your representation.
    Concretely, if your odd rep is $\gamma^{(2k + 1)}$, then your even rep 
    is $\{ \gamma^{(2k + 1)} \otimes \sigma^a, I_{2k} \otimes \sigma^b \}$.
\end{itemize}


\end{document}
