% =====================================================================
% Appendix: conventions
% Grows as chapters need it. Started with chapter I.1 (Wick rotation,
% topological terms in Euclidean signature).
% =====================================================================
\section{Conventions}
\label{app:conventions}

\paragraph{Units and signature.} We set $\hbar = c = 1$ throughout. The Lorentzian metric signature is \textbf{mostly-plus}, $\eta_{\mu\nu} = \mathrm{diag}(-,+,+,+)$ (and its obvious truncations in other dimensions). Consequently a free scalar has
\begin{equation}
\el = -\tfrac{1}{2}\,\partial_\mu \phi\, \partial^\mu \phi - \tfrac{1}{2} m^2 \phi^2 ,
\qquad p^2 = p_\mu p^\mu = -m^2 \ \text{on shell}.
\end{equation}
Srednicki and Weinberg use this signature; Peskin--Schroeder does not. Much of the symmetry and CFT literature works directly in Euclidean signature, where the issue is moot.

\paragraph{Wick rotation.} Euclidean time is $t = -i\tau$ with $\tau$ real, so that $e^{iS_{\rm L}} = e^{-S_{\rm E}}$ and a standard kinetic term rotates as
\begin{equation}
i \int dt\, \tfrac{1}{2} \dot q^2 \;\longrightarrow\; - \int d\tau\, \tfrac{1}{2} \left(\frac{dq}{d\tau}\right)^{\!2},
\end{equation}
i.e.\ $S_{\rm E}$ is positive for a standard kinetic term. Thermal physics at temperature $T = 1/\beta$ lives on a Euclidean time circle of circumference $\beta$: $Z(\beta) = \tr e^{-\beta H}$.

\paragraph{Topological terms stay imaginary.} A term in the action with no dependence on the metric acquires no factor of $i$ under Wick rotation: it enters the Euclidean weight as a pure phase. For the $\theta$-term of quantum mechanics on a circle (chapter~I.1),
\begin{equation}
i \,\frac{\theta}{2\pi} \int dt\, \dot q \;\longrightarrow\; i\, \frac{\theta}{2\pi} \oint dq = i\theta w, \qquad w \in \ints,
\end{equation}
and the Euclidean weight is $e^{-S_{\rm E} + i\theta w}$. Forgetting this $i$ (i.e.\ writing $e^{-\theta w}$) is a classic error; the same rule governs $\theta F\tilde F$ in 4d and Chern--Simons terms in 3d.

\paragraph{Euclidean operators and conjugation (from chapter I.2).} Euclidean Heisenberg evolution is $\op(\tau, \mathbf{x}) = e^{H\tau}\,\op(0,\mathbf{x})\,e^{-H\tau}$, so a Lorentzian-Hermitian scalar satisfies $\op(\tau,\mathbf{x})^\dagger = \op(-\tau, \mathbf{x})$: Euclidean conjugation reflects through the quantization surface. Operators with spin additionally pick up a factor $-i$ per Lorentzian time index under Wick rotation ($\op_{\rm E}^{0\cdots} = -i\,\op_{\rm L}^{0\cdots}$, etc.), so that Euclidean correlators are $SO(d)$-covariant; conjugation then carries a time-reflection matrix per index. See Simmons-Duffin \arxiv{1602.07982}, \S7.1.

\paragraph{Radial quantization conventions (from chapter I.2).} Following Simmons-Duffin \arxiv{1602.07982}: Euclidean signature; the dilatation generator $D$ is Hermitian with $[D, \op(0)] = \Delta\,\op(0)$, $[P_\mu, \op(x)] = \partial_\mu \op(x)$, and $[K_\mu, \op(0)] = 0$ for a primary; algebra $[D, P_\mu] = P_\mu$, $[D, K_\mu] = -K_\mu$, $[K_\mu, P_\nu] = 2\delta_{\mu\nu}D - 2M_{\mu\nu}$. Conjugation in radial quantization is inversion: $\op(x)^{\dagger} = x^{-2\Delta}\,\op(x^\mu/x^2)$ for a Hermitian scalar, whence $P_\mu^\dagger = K_\mu$, $D^\dagger = D$.

\paragraph{Stress tensor (from chapter I.2).} Euclidean: $T_{\mu\nu} = \frac{2}{\sqrt g}\frac{\delta S_{\rm E}}{\delta g^{\mu\nu}}$; Lorentzian (mostly-plus): $T_{\mu\nu} = -\frac{2}{\sqrt{-g}}\frac{\delta S}{\delta g^{\mu\nu}}$, giving $T_{00} \ge 0$ for the standard scalar. Ward identity with our signs:
\begin{equation}
\partial_\mu \langle T^{\mu\nu}(x)\,\op_1(x_1)\cdots\rangle = -\sum_i \delta^d(x - x_i)\,\partial^\nu_{x_i}\langle \op_1(x_1)\cdots\rangle .
\end{equation}
Improvements $T_{\mu\nu} \to T_{\mu\nu} + \xi(\delta_{\mu\nu}\partial^2 - \partial_\mu\partial_\nu)\op$ are identically conserved and change no charges; ``traceless'' always means ``for the improved representative.''

\paragraph{Symmetry and CFT conventions.} When the chapters reach the symmetry material (Part II) we follow the conventions of Shao's TASI lectures \arxiv{2308.00747}; for CFT material (Part III), Simmons-Duffin \arxiv{1602.07982}. Conventions for $p$-form gauge fields, flux normalization, and Dirac matrices will be pinned here when first needed.
